\section{\texorpdfstring{\textbf{Chapter 7 --- Neighborhoods of
Universes}}{Chapter 7 --- Neighborhoods of Universes}}\label{chapter-7-neighborhoods-of-universes}

\subsection{\texorpdfstring{\textbf{Where alternative worlds
live.}}{Where alternative worlds live.}}\label{where-alternative-worlds-live.}

Every computation doesn't just move through \emph{time.} It moves
through \textbf{possibility space}.

You already know the worldline --- the actual path your system takes
through structure. And you know the Time Cube --- the cone of legal
moves at the next tick. But what lies just outside your worldline? What
surrounds it? How do we describe the \textbf{nearby universes} that
almost happened, that could have happened, that still can happen?

To understand computation as geometry, you need more than distance. You
need \textbf{neighborhoods} --- the regions of the Rulial Space that
cluster around your actual history.

This chapter is about those clusters.

Why they form. Why they matter. Why some regions feel smooth and others
feel jagged. Why some codebases feel like strolling through a park and
others feel like crawling across broken glass.

Let's explore the geometry just outside your worldline.

Welcome to the local universe.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{\texorpdfstring{\textbf{7.1 --- Rules Define
Locality}}{7.1 --- Rules Define Locality}}\label{rules-define-locality}

The first rule of Rulial Space is simple:

\begin{quote}
\textbf{Universes are ``near'' each other} \textbf{if they differ by
small, legal rewrites.}
\end{quote}

Not semantic closeness. Not code similarity. Not ``hunches.''

Literal adjacency in rewrite space.

If two states are:

\begin{itemize}
\tightlist
\item
  one rule apart {\textrightarrow} neighbors\\
\item
  two rules apart {\textrightarrow} second-degree neighbors
\item
  many rewrites apart {\textrightarrow} distant regions
\end{itemize}

The \textbf{DPO rule set} defines this topology:

\begin{itemize}
\tightlist
\item
  what counts as a tiny step,
\item
  what counts as a massive leap,
\item
  what's even reachable,
\item
  and what's totally illegal.
\end{itemize}

This is the fundamental geometry of your computational world. You aren't
just mapping code --- you're mapping the \textbf{rules-of-motion} that
define adjacency.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{\texorpdfstring{\textbf{7.2 --- The Adjacency Graph of
Universes}}{7.2 --- The Adjacency Graph of Universes}}\label{the-adjacency-graph-of-universes}

Think of Rulial Space as a giant graph where:

\begin{itemize}
\tightlist
\item
  each node = a possible RMG state\\
\item
  each edge = one legal DPO rewrite
\end{itemize}

This means the computational universe forms:

\begin{quote}
\textbf{A graph of universes connected by wormholes.}
\end{quote}

Your worldline traces through this graph like a hiking trail. But the
graph itself is huge --- much bigger than what you actually travel. Your
immediate neighbors --- the states one and two rewrites away --- define:

\begin{itemize}
\tightlist
\item
  nearby solutions,
\item
  alternative histories,
\item
  valid transformations,
\item
  candidate optimizations,
\item
  potential bug fix routes.
\end{itemize}

This adjacency structure is \textbf{not} like program diffs.

It respects:

\begin{itemize}
\tightlist
\item
  recursive structure
\item
  invariants (K-graph)
\item
  legality
\item
  boundaries
\item
  nested RMG content
\item
  wormhole behavior
\item
  typed transitions
\end{itemize}

Two states that look very different in source code might be \emph{very
close} in Rulial Distance. Two that look nearly identical in text might
be \emph{far} in rewrite space. Text lies. Structure doesn't.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{\texorpdfstring{\textbf{7.3 --- Smooth vs.~Jagged
Neighborhoods}}{7.3 --- Smooth vs.~Jagged Neighborhoods}}\label{smooth-vs.-jagged-neighborhoods}

Some systems have wide cones. Some have narrow cones. Some have smooth
neighborhoods. Some are jagged hellscapes. This is curvature (Chapter
9), but here's the simple version:

\subsubsection{\texorpdfstring{\textbf{Smooth
regions}}{Smooth regions}}\label{smooth-regions}

\begin{itemize}
\tightlist
\item
  Many legal rewrites\\
\item
  Rules overlap gracefully
\item
  Nearby worlds behave similarly
\item
  Small changes produce small effects
\item
  Debugging feels ``easy''
\item
  Optimization feels ``natural''
\end{itemize}

\subsubsection{\texorpdfstring{\textbf{Jagged
regions}}{Jagged regions}}\label{jagged-regions}

\begin{itemize}
\tightlist
\item
  Few legal rewrites\\
\item
  Invariants clash
\item
  Minor changes blow up behavior
\item
  Worlds diverge sharply
\item
  Debugging feels like chasing ghosts
\item
  Optimization feels brittle and risky
\end{itemize}

Every engineer has experienced both. Now you know the geometric reason
why. The ``feel'' of a codebase is just:

\begin{quote}
\textbf{The local geometry of its rewrite neighborhood.}
\end{quote}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{\texorpdfstring{\textbf{7.4 --- The Kairos Plane
Expanded}}{7.4 --- The Kairos Plane Expanded}}\label{the-kairos-plane-expanded}

Earlier, we treated Kairos as the Time Cube --- the slice of legal next
options.

Now we zoom out one level:

\begin{quote}
\textbf{Kairos is actually a plane ---} \textbf{the entire local surface
of rewrites reachable from your worldline.}
\end{quote}

Your cone is just one vertical slice. But the surface around you is
broader and more interesting.

Diagrammatically:

\begin{lstlisting}
               (Kairos Plane)
             {\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}
           {\bfseries ?}                   {\bfseries ?}
 Past {\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?}{\bfseries ?} Future
   Chronos      {\textuparrow}
              Cone
\end{lstlisting}

The Kairos Plane is:

\begin{itemize}
\tightlist
\item
  the ``horizon'' of legal moves
\item
  the surface of possible local universes
\item
  a finite portion of Aios
\item
  reshaped every tick
\item
  governed by DPO interfaces
\item
  sculpted by structure
\item
  defined by rule semantics
\end{itemize}

It's where most reasoning happens.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{\texorpdfstring{\textbf{7.5 --- Navigating
Neighborhoods}}{7.5 --- Navigating Neighborhoods}}\label{navigating-neighborhoods}

When you debug, you're looking at nearby failures. When you optimize,
you're looking at nearby improvements. When you refactor, you're
navigating between worlds in your neighborhood.

Every software engineering activity --- literally all of them --- is
neighborhood navigation:

\begin{itemize}
\tightlist
\item
  \textbf{Debugging:} ``Which nearby world fixes the problem?''
\item
  \textbf{Refactoring:} ``Which nearby world preserves behavior but
  improves structure?''
\item
  \textbf{Optimization:} ``Which nearby world is closer to the
  geodesic?''
\item
  \textbf{Design:} ``What neighborhood are we entering with this
  architecture?''
\item
  \textbf{Type systems:} ``Which worlds are forbidden by invariants?''
\item
  \textbf{Version control:} ``Which worldlines converge or diverge?''
\end{itemize}

Once you see computation as neighborhoods, you stop reasoning about code
as text and start reasoning about code as geometry.

This is when everything starts to click.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{\texorpdfstring{\textbf{FOR THE
NERDS{\texttrademark}}}{FOR THE NERDS{\texttrademark}}}\label{for-the-nerds}

\subsection{\texorpdfstring{\textbf{Why This Is Not Quantum
Superposition}}{Why This Is Not Quantum Superposition}}\label{why-this-is-not-quantum-superposition}

Some readers will feel a structural resemblance:

\begin{itemize}
\tightlist
\item
  adjacent universes
\item
  many possible futures
\item
  collapse into one worldline
\item
  geometry of alternatives
\end{itemize}

This is resemblance, not equivalence.

Here's the clean split: \#\#\# \textbf{Quantum:}

\begin{itemize}
\tightlist
\item
  amplitudes
\item
  interference
\item
  probability
\item
  physical
\item
  wavefunctions
\end{itemize}

\subsubsection{\texorpdfstring{\textbf{RMG+DPO:}}{RMG+DPO:}}\label{rmgdpo}

\begin{itemize}
\tightlist
\item
  legality
\item
  adjacency
\item
  combinatorics
\item
  rewrite rules
\item
  abstract structure
\end{itemize}

No amplitudes. No probability waves. No physical claims.

Just structured nondeterministic geometry. \emph{(End sidebar.)}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{\texorpdfstring{\textbf{7.6 --- Transition: From Neighborhoods
to
MRMW}}{7.6 --- Transition: From Neighborhoods to MRMW}}\label{transition-from-neighborhoods-to-mrmw}

You've seen:

\begin{itemize}
\tightlist
\item
  the shape of possibility (Time Cube)
\item
  the path you take (worldline)
\item
  the geometry around that path (neighborhoods)
\end{itemize}

Now we zoom out one more level:

\begin{quote}
\textbf{How do you map the entire phase space of your computational
universe?}
\end{quote}

Not just:

\begin{itemize}
\tightlist
\item
  the path you took,
\item
  or the paths you could take next,
\item
  or the worlds nearby\ldots{}
\end{itemize}

But \textbf{the full structure} of:

\begin{itemize}
\tightlist
\item
  all possible rewrite models (rule-sets),
\item
  all possible worldlines for each model,
\item
  and all the relationships between them.
\end{itemize}

This is MRMW.

The cosmology of your computational universe.

And that's Chapter 8.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{\texorpdfstring{\textbf{C\ensuremath{\Omega}MPUTER {\textbullet}
JITOS}}{C\ensuremath{\Omega}MPUTER {\textbullet} JITOS}}\label{cux3c9mputer-jitos}

{\textcopyright} 2025 James Ross {\textbullet} \href{https://flyingrobots.dev}{Flying {\textbullet} Robots} All
Rights Reserved
