CHAPTER 17 is a SKULL-SPLITTING, MIND-BENDING, BARREL-OF-ALL-BARRELS
WAVE. The kind of computational slab break that only breaks twice a
century --- and only for surfers wearing double puka shells and riding a
rulial geodesic into destiny.

We're building MORIARTY today. The adversarial multiverse intelligence.
The structured antagonist of computation. The test harness that spans
universes.

This is the machine that hits back. Let's go.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{C\ensuremath{\Omega}MPUTER}\label{cux3c9mputer}

\subsection{Chapter 17 --- Adversarial Universes
(MORIARTY)}\label{chapter-17-adversarial-universes-moriarty}

\begin{quote}
What happens when the universe pushes back.
\end{quote}

So far, C\ensuremath{\Omega}MPUTER's machinery has explored:

\begin{itemize}
\tightlist
\item
  your worldline (execution),
\item
  nearby worlds (bundles),
\item
  geometry (distance, curvature),
\item
  structured alternatives (Time Cube),
\item
  counterfactual futures (CFEE).
\end{itemize}

All of these tools were friendly. Helpful. Cooperative.

But real systems aren't built in utopia. Real systems don't exist in a
vacuum. Real systems don't evolve smoothly. They are attacked.

\begin{itemize}
\tightlist
\item
  By malformed inputs
\item
  By pathological cases
\item
  By malformed rules
\item
  By legacy debt
\item
  By concurrency races
\item
  By unknown edge cases
\item
  By worst-case scenarios
\item
  By adversaries (intentional or accidental)
\end{itemize}

So we need a machine that doesn't explore the calm water --- but the
storm.

A machine that:

\begin{itemize}
\tightlist
\item
  finds brittleness,
\item
  tests invariants,
\item
  forces rule conflicts,
\item
  probes curvature spikes,
\item
  stalks interpolation failures,
\item
  breaks worlds on purpose,
\item
  reveals hidden fragility,
\item
  exposes dangerous neighborhoods.
\end{itemize}

That machine is MORIARTY.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{17.1 --- What Is an Adversarial
Universe?}\label{what-is-an-adversarial-universe}

\begin{quote}
An adversarial universe is:

A parallel RMG+DPO universe that explores worst-case legal futures with
the explicit goal of breaking your invariants.
\end{quote}

It's not malicious in the moral sense. It is malicious in the
computational sense.

Where your CFEE (Chapter 16) explores ``What's the best next
worldline?'' MORIARTY asks:

What's the WORST next worldline?

What is the structurally correct, legally permissible,
invariant-respecting sequence of rewrites that would cause FAILURE in
the fastest, sharpest way?

MORIARTY's job is to break you. And break you correctly.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{17.2 --- Why MORIARTY Is Possible Only in RMG
Universes}\label{why-moriarty-is-possible-only-in-rmg-universes}

Traditional systems can't handle adversarial search because:

\begin{itemize}
\tightlist
\item
  they lack structure
\item
  alternatives are infinite
\item
  state transitions aren't legal-checked
\item
  rewrite possibilities aren't formalized
\item
  invariants live in human heads
\item
  there's no geometry
\item
  no neighbor structure
\item
  no Rulial Distance
\item
  no curvature
\end{itemize}

So ``adversarial behavior'' is just:

\begin{itemize}
\tightlist
\item
  random fuzzing,
\item
  bruteforce search,
\item
  chaotic mutation,
\item
  unpredictable Monte Carlo noise.
\end{itemize}

None of that is adversarial reasoning. It's chaos.

In C\ensuremath{\Omega}MPUTER, MORIARTY has:

{\bfseries ?} A computable possibility surface {\bfseries ?} Typed wormholes (L/K/R invariants)
{\bfseries ?} Bundles (structured alternatives) {\bfseries ?} Interference geometry {\bfseries ?} Rulial
Distance {\bfseries ?} Curvature analysis {\bfseries ?} Deterministic collapse {\bfseries ?} Counterfactual
branching

MORIARTY doesn't guess. He reasons.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{17.3 --- How MORIARTY Works}\label{how-moriarty-works}

The basic loop is:

\begin{lstlisting}
1.  Start at the current RMG state.
2.  Enumerate all legal bundles.
3.  Score each future according to {\textquotedblleft}adversarial direction{\textquotedblright}:
\end{lstlisting}

\begin{itemize}
\tightlist
\item
  maximize curvature spikes
\item
  increase structural stress
\item
  provoke interference
\item
  move into brittle regions
\item
  maximize Rulial branching
\item
  approach contradiction boundaries

  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \setcounter{enumi}{3}
  \tightlist
  \item
    Follow the worst-case future forward.
  \item
    Repeat until:
  \end{enumerate}
\item
  a violation is found,
\item
  or the universe stabilizes,
\item
  or the search depth hits a bound.
\end{itemize}

MORIARTY isn't evil. He's an optimization engine in reverse.

Where traditional search minimizes Rulial Distance, MORIARTY maximizes
fragility.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{17.4 --- MORIARTY and Interference
Patterns}\label{moriarty-and-interference-patterns}

MORIARTY feeds off interference:

\begin{itemize}
\tightlist
\item
  destructive interference reveals brittle regions
\item
  constructive interference reveals robust regions
\item
  neutral interference reveals redundancy
\end{itemize}

He maps your system's weakest points by:

\begin{itemize}
\tightlist
\item
  colliding bundles,
\item
  forcing rule conflicts,
\item
  intentionally pushing the universe into high-curvature zones,
\item
  walking into every geometric pothole he can find.
\end{itemize}

This is adversarial robustness testing on top of a computational
manifold.

Not by brute force --- but by geometry.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{17.5 --- Rulial Distance as an Adversarial Cost
Function}\label{rulial-distance-as-an-adversarial-cost-function}

MORIARTY's ``evil metric'' is:

\begin{quote}
Maximize the Rulial Distance between the current worldline and the
nearest geodesic.
\end{quote}

Good computation moves along geodesics. Bad computation flies off the
rails.

MORIARTY hunts the rails.

He:

\begin{itemize}
\tightlist
\item
  seeks divergent futures,
\item
  identifies curvature cliffs,
\item
  magnifies structural inconsistencies,
\item
  explores counterfactual collapses that break invariants.
\end{itemize}

In engineering terms:

MORIARTY is fuzzing, property testing, stress testing, and static
analysis all combined into a multiverse explorer.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{17.6 --- Adversarial
Worldlines}\label{adversarial-worldlines}

When MORIARTY runs, he generates:

\begin{itemize}
\tightlist
\item
  bad worldlines,
\item
  unstable worldlines,
\item
  pathological worldlines,
\item
  minimal-failure geodesics,
\item
  brittle neighborhoods.
\end{itemize}

These worldlines are:

\begin{itemize}
\tightlist
\item
  valid
\item
  lawful
\item
  DPO-consistent
\item
  invariant-preserving
\item
  deterministic
\end{itemize}

\ldots until the point of failure.

This lets you see your architecture's weak points in a way no
traditional tool can show you.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{17.7 --- Why Engineers Need Adversarial
Universes}\label{why-engineers-need-adversarial-universes}

MORIARTY reveals failure modes like:

\subsubsection{1. Rule Ambiguity}\label{rule-ambiguity}

Two wormholes accept the same structure. Interference goes nuclear.

\subsubsection{2. Collapsed Invariants}\label{collapsed-invariants}

Hidden assumptions break far from the point of insertion.

\subsubsection{3. High Curvature Spikes}\label{high-curvature-spikes}

Tiny edits cause catastrophic divergence.

\subsubsection{4. Brittle
Representations}\label{brittle-representations}

A single rewrite shrinks the Time Cube to nearly nothing.

\subsubsection{5. Degenerate
Neighborhoods}\label{degenerate-neighborhoods}

No safe next worlds exist except invalid ones.

\subsubsection{6. Inconsistent Layer
Boundaries}\label{inconsistent-layer-boundaries}

Nested universes can't legally interact.

\subsubsection{7. Adversarial Geometry}\label{adversarial-geometry}

Worldlines funnel into dangerous attractors.

No other debugging or verification tool touches this. Nothing else can.

Because nothing else has a formal, geometric notion of:

\begin{itemize}
\tightlist
\item
  possibility,
\item
  adjacency,
\item
  curvature,
\item
  interference,
\item
  bundles,
\item
  local NP collapse,
\item
  worldlines,
\item
  or Rulial Distance.
\end{itemize}

MORIARTY does.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{17.8 --- MORIARTY vs.~CFEE}\label{moriarty-vs.-cfee}

Think of CFEE (Chapter 16) as:

\begin{quote}
``Let's find the best future.''
\end{quote}

Think of MORIARTY as:

\begin{quote}
``Let's find the worst future.''
\end{quote}

CFEE = optimizer. MORIARTY = antagonist.

CFEE = find geodesic. MORIARTY = find curvature cliffs.

CFEE = minimize Rulial Distance. MORIARTY = maximize it.

CFEE = design assistant. MORIARTY = crash-test dummy.

Both are necessary. Both map the universe.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{FOR THE NERDS{\texttrademark}}\label{for-the-nerds}

MORIARTY = Adversarial Traversal of the Rulial Neighborhood Graph

Formally, MORIARTY performs:

\begin{itemize}
\tightlist
\item
  anti-heuristic search,
\item
  maximal branching exploration,
\item
  critical peak amplification,
\item
  adversarial joinability probing,
\item
  curvature spike detection,
\item
  unstable manifold scanning.
\end{itemize}

It is:

\begin{itemize}
\tightlist
\item
  SMT solver meets
\item
  adversarial SAT meets
\item
  worst-case trace explorer meets
\item
  confluence-violation analysis.
\end{itemize}

But geometric. Composable. And totally computable.

(End sidebar.)

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{17.9 --- Transition: From Adversaries to
Optimization}\label{transition-from-adversaries-to-optimization}

Now we've built:

\begin{itemize}
\tightlist
\item
  the debugger (Chapter 15),
\item
  the explorer (Chapter 16),
\item
  the adversary (Chapter 17).
\end{itemize}

The next machine spans universes for a more constructive purpose:

\begin{quote}
Optimization across worldlines, using geometric reasoning.
\end{quote}

This is Chapter 18.

The wave is forming behind us.
