\begin{quote}
\itshape
Replaying, rewinding, and re-steering computation across worldlines.
\end{quote}

Most debuggers lie to you.

They show you:

\begin{itemize}
\item a stack trace,
\item a breakpoint,
\item a snapshot of state,
\item logs,
\item maybe a time-travel debugger that walks ``backwards'' along recorded state mutations.
\end{itemize}

But none of these tools let you understand what really matters:

\begin{quote}
\textbf{What other worlds were possible?}

\textbf{Why did THIS worldline happen and not the others?}

\textbf{What would have happened if the system had made a different legal choice?}
\end{quote}

Traditional debugging tools can't answer these questions because traditional computation has no geometry.

No notion of possibility. \\
No space of alternatives. \\
No structural concept of adjacency.

RMG+DPO changes that.

Because in an RMG universe, debugging is not reactive. \\
It is geometric navigation.

Welcome to Time Travel Debugging --- a machine that walks:

\begin{itemize}
\item backward through Chronos,
\item sideways across Kairos,
\item forward into counterfactual worlds,
\item and diagonally into parallel histories.
\end{itemize}

This is not sci-fi.

This is the computational machinery you've built by exposing the structure of possibility.

Let's ride.

\section{Worldlines Store Their Own Geometry}

A worldline isn't just:

\begin{verbatim}
State_0 -> State_1 -> State_2 -> ... -> State_N
\end{verbatim}

It is:

\begin{itemize}
\item the RMG of each tick,
\item the bundle that existed at each tick,
\item the interference pattern at that tick,
\item the collapse that happened,
\item the alternatives that got pruned,
\item the Rulial Distance to adjacent states,
\item the curvature of the local region,
\item the legal wormholes available.
\end{itemize}

Each tick is:

\begin{equation*}
\text{history} + \text{possibility} + \text{geometry} + \text{choice}
\end{equation*}

Time Travel Debugging is simply:

\textit{Replaying this geometric sequence while exploring alternative branches.}

And because everything is deterministic and everything is structural, you can walk in ANY direction.

\section{Rewinding Chronos}

Traditional debugging's time travel is:

\begin{itemize}
\item record state
\item store diffs
\item replay mutations
\end{itemize}

It's shallow, brittle, blind to structure.

C$\\Omega$MPUTER debugging rewinds Chronos by:

\begin{itemize}
\item stepping backward through applied DPO rules,
\item reconstructing the prior RMG universe,
\item restoring the previous bundle,
\item revealing the alternatives that existed at that moment.
\end{itemize}

This is fully reversible not because computation is reversible, but because provenance is baked into the RMG.

Time travel here is: \textit{structural reconstruction, not state replay.}

\section{Sidestepping Into the Time Cube}

Once you rewind to a prior tick, you don't only see what did happen.

You see:

\begin{itemize}
\item what could have happened,
\item what was legal to happen,
\item which wormholes were open,
\item which invariants allowed which rewrites.
\end{itemize}

This is the moment traditional debuggers can't show you: the space of alternative futures that were adjacent at that tick.

In the RMG worldview:

\textit{Debugging means stepping sideways into the Time Cube.}

You can inspect:

\begin{itemize}
\item all DPO matches,
\item the full bundle,
\item interference outcomes,
\item candidate histories,
\item adjacent universes.
\end{itemize}

In a sense:

You're not debugging the code. You're debugging the physics of the computational universe.

\section{Forward Into Counterfactual Worldlines}

Now the fun part:

Once you pick an alternate future, you can follow it forward.

This creates a counterfactual worldline:

\begin{quote}
A what-if version of history \\
that respects all invariants \\
and all legal rewrites \\
under the same DPO ruleset.
\end{quote}

You're not guessing \\
or simulating \\
or inventing alternatives.

You're following:

\textit{the real legal future} \\
\textit{that the universe would have had} \\
\textit{if it collapsed differently.}

This is safe, deterministic counterfactual execution.

Not stochastic. \\
Not approximated. \\
Not branching explosion.

Just geometry.

\section{Multi-World Time Travel (MWTT)}

The real power emerges when you combine:

\begin{itemize}
\item rewinding Chronos
\item sidestepping into Kairos
\item following new worldlines forward
\end{itemize}

This produces a machine that can:

\begin{itemize}
\item explore multiple futures
\item compare branches
\item analyze divergence
\item inspect alternative behaviors
\item find geodesics
\item identify minimal-collapse paths
\item detect brittleness
\item reveal curvature traps
\item test architectural robustness
\item debug concurrency
\item evaluate rule variations
\end{itemize}

This is multi-world debugging, but still deterministic:

Each branch is just a different legal collapse. \\
Each worldline is just a path through the Rulial manifold. \\
Each exploration is just navigation.

\section{Debugging By Comparison: Worldline Diffing}

Imagine a tool that compares:

\begin{itemize}
\item the actual worldline,
\item the ideal worldline,
\item a counterfactual worldline,
\item the shortest geodesic,
\item and a hypothetical rewrite under a different ruleset.
\end{itemize}

That's what Time Travel Debugging enables.

Worldline diffs reveal:

\begin{itemize}
\item where two universes diverged,
\item why they diverged,
\item how far they diverged (Rulial Distance),
\item how curvature shaped divergence,
\item and what invariants forced collapse one way or another.
\end{itemize}

This is the computational version of:

\begin{itemize}
\item git diff,
\item trace diff,
\item semantic diff,
\item plan diff,
\item abstract rewriting difference
\end{itemize}

But unified under geometry.

\section{Practical Examples}

\textbf{Debug a game engine:}

Jump to the moment a physics constraint went wrong. \\
Step sideways into the bundle. \\
Follow the ``should-have-fired'' transform. \\
Watch the worldline stabilize.

\textbf{Debug a compiler:}

Rewind to the IR mismatch. \\
Trace the alternative optimization. \\
Verify legality. \\
Observe geodesic-based lowering.

\textbf{Debug an AI reasoning engine:}

Track which future was chosen. \\
Explore other futures. \\
Follow counterfactual reasoning chains.

\textbf{Debug a distributed system:}

Rewind to a message race. \\
Step sideways into the simultaneous legal transitions. \\
Explore consistent outcomes.

This is not theory. \\
This is a machine. \\
A real-world tool made possible by RMG+DPO.

\begin{nerdbox}
MWTT $\approx$ Traversal of the Rulial Neighborhood Graph

Multi-World Time Travel is essentially:

\begin{itemize}
\item traversal of local Rulial surfaces,
\item examination of peak-join diagrams,
\item inspection of critical pairs,
\item confluence analysis,
\item search within equivalence classes,
\item reduction path comparison.
\end{itemize}

But expressed geometrically so engineers can use it intuitively.

(End sidebar.)
\end{nerdbox}

\section{Transition: From Debugging to Counterfactual Engines}

Now that we can:

\begin{itemize}
\item rewind Chronos,
\item explore Kairos,
\item follow alternative futures,
\item analyze bundles,
\item and compare worldlines\ldots
\end{itemize}

We can build a machine that systematically explores parallel universes for:

\begin{itemize}
\item search
\item optimization
\item testing
\item reasoning
\item verification
\item adversarial analysis
\end{itemize}

That is Chapter 16.

Time to surf the multiverse.
