\section{JITOS}\label{jitos}

\begin{quote}
\emph{JITOS is the operating system for C\ensuremath{\Omega}MPUTER.}
\end{quote}

\subsection{What is C\ensuremath{\Omega}MPUTER?}\label{what-is-cux3c9mputer}

\textbf{C\ensuremath{\Omega}MPUTER} is a computational model based on deterministic graph
rewrites applied to a \textbf{Recursive Metagraph} (RMG) with
holographic properties.

The system updates via a scheduler that applies rules to the RMG
whenever specific conditions are met. Because the nodes are holographic,
every calculated value carries its entire provenance.

This makes C\ensuremath{\Omega}MPUTER a time machine. You can reconstruct the history of
any value---stepping backward from the moment it was calculated, to when
the program started, to the build process, and finally to the moment the
source code was written. Each event is a graph transformation encoded in
the value's provenance chain: the causal graph.

\subsubsection{The Worldline}\label{the-worldline}

The basic unit of the C\ensuremath{\Omega}MPUTER is the holographic node. These nodes form
a causal DAG (Directed Acyclic Graph)---an append-only ``worldline''
that represents the computer's motion through AI\ensuremath{\Omega}N, the geometric
spacetime of computing. Because every node encodes its causal history,
the system is cryptographically tamper-evident. Because it is
append-only, it is fully auditable. C\ensuremath{\Omega}MPUTER is a glass box.

\subsubsection{Deterministic State}\label{deterministic-state}

The causal graph is an immutable ledger of rewrites. Given the same
initial state and the same sequence of rewrites, the resulting graph
will be isomorphically identical every time, down to the bit.

\subsection{What's JITOS?}\label{whats-jitos}

JITOS is the operating system for C\ensuremath{\Omega}MPUTER.

Its primary function is to bridge the gap between human workflow and
machine reality via Holographic Projection.

\begin{itemize}
\tightlist
\item
  \textbf{For Humans (The Projection):} JITOS materializes views of the
  causal DAG that look and behave exactly like a standard filesystem.
  Humans see files; IDEs see folders; compilers see source trees. But
  these are just transient projections---an interface layer generated
  on-the-fly from the underlying graph.
\item
  \textbf{For Agents (The Reality):} AI agents bypass the projection
  entirely. They do not need to parse linear text files or navigate
  directories. Instead, they interact directly with the causal DAG,
  manipulating the raw graph structure, dependencies, and provenance
  chains.
\end{itemize}

There are no processes. Instead, agency occurs in shadow working
sets---isolated, parallel branches of the universe held in
superposition.

JITOS is the interface. For humans, it simulates the familiar desktop.
For AIs, it grants direct access to the physics of the system.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{\texorpdfstring{\textbf{C\ensuremath{\Omega}MPUTER {\textbullet}
JITOS}}{C\ensuremath{\Omega}MPUTER {\textbullet} JITOS}}\label{cux3c9mputer-jitos}

{\textcopyright} 2025 James Ross {\textbullet} \href{https://flyingrobots.dev}{Flying {\textbullet} Robots} All
Rights Reserved
