\section{RFC-0021}\label{rfc-0021}

JIT RFC-0021

Multi-Universe Federation Protocol (MUFP v1.0)

Interoperability, Intercosmic Sync, and Cross-Universe Provenance Flow

Status: Draft\\
Author: James Ross\\
Contributors: JIT Community\\
Requires:\\
- RFC-0011 Distributed Sync\\
- RFC-0017 Agent Identity\\
- RFC-0020 Security Model\\
Start Date: 2025-11-30\\
Target Spec: JITOS v0.x\\
License: TBD**

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Summary
\end{enumerate}

JIT supports:\\
- sync within one repository\\
- causal consistency within one DAG\\
- one universe's nodes

But the next frontier is:

multiple universes

each with their own:\\
- separate DAG\\
- separate causal fabric\\
- separate refs\\
- separate SWS\\
- separate identity graphs\\
- separate provenance\\
- separate storage tiers

Federation enables:\\
- multi-repo reasoning\\
- cross-universe linking\\
- inter-cosmic provenance\\
- DAG-to-DAG references\\
- global build systems\\
- multi-project dependency graphs\\
- external truth linking\\
- ``universe imports''\\
- C\ensuremath{\Omega}MPUTER-level world fusion

This RFC defines how universes recognize each other, communicate,
exchange causality subsets, and maintain containment without collapse.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  Motivation
\end{enumerate}

The world is not a single repo.\\
Software ecosystems are not single universes.\\
Scientific computation spans multiple DAGs.\\
Simulation engines create parallel spaces.\\
Agents must reason across multiple knowledge graphs.\\
Organizations maintain separate worlds for security.

We must define:\\
- how these universes SEE each other\\
- how they link\\
- how they exchange truth\\
- how they maintain separation\\
- how they sync without collapsing\\
- how they track provenance across cosmos boundaries

This is the Inter-DAG Protocol.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  Universe Identity
\end{enumerate}

Every universe MUST have an identity:

UniverseID = \{\\
id: uuid,\\
public\_key: bytes,\\
metadata: \{\\
name: string,\\
owner: string,\\
description: string,\\
creation\_ts: u64\\
\}\\
\}

This is like Agent Identity but scaled up.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  Universe Boundaries
\end{enumerate}

A universe boundary is defined by:\\
- root nodes\\
- topological closure\\
- storage domain\\
- provenance domain\\
- identity domain\\
- ref set

No universe may mutate another.

Universes are disjoint DAGs\\
that MAY link through federation edges.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  Federation Edges
\end{enumerate}

A federation edge is a special node:

FederationEdge \{\\
type: ``federation-edge'',\\
local\_node: NodeID,\\
remote\_universe: UniverseID,\\
remote\_node: RemoteNodeID,\\
metadata: \{\\
purpose: string,\\
trust\_level: string,\\
timestamps: \{\ldots\}\\
\}\\
\}

Purpose examples:\\
- dependency\\
- import\\
- reference\\
- causal linkage\\
- provenance attribution\\
- simulation embedding

These edges do NOT merge DAGs.\\
They create inter-universe links.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{5}
\tightlist
\item
  Federation Sync
\end{enumerate}

Federation sync is:\\
- NOT full sync\\
- NOT merging histories\\
- NOT rewriting\\
- NOT unifying universes

Federation sync is:\\
- capability negotiation\\
- frontier announcement\\
- provenance sharing\\
- partial subgraph import\\
- dependency verification\\
- policy-driven trust

It is Inter-OS sync, not repo sync.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{6}
\tightlist
\item
  Federation Invariants
\end{enumerate}

Invariant 1---No DAG Fusion

Federated DAGs MUST remain separate.\\
No merge across universes.

Invariant 2---Unidirectional Links

Federation edges MUST declare direction.\\
Local $\rightarrow$ remote\\
Remote $\rightarrow$ local\\
or both, but explicit.

Invariant 3---Trust Levels

Each universe MUST declare:

trust\_level \ensuremath{\in} \{trusted, semi-trusted, untrusted\}

Invariant 4---No Causal Pollution

Nodes from foreign universes MUST NOT be treated as local nodes.

Invariant 5---Deterministic Import

Imported nodes MUST remain exactly identical to their origin.

Invariant 6---Provenance Persistence

Cross-universe provenance MUST store both UniverseIDs.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{7}
\tightlist
\item
  Federation Sync Stages
\end{enumerate}

Stage 1---``Hello, Universe.''

Universes exchange:\\
- UniverseID\\
- capabilities\\
- policy rules\\
- public keys\\
- feature-sets\\
- version

Stage 2---Capability Intersect

Intersect:\\
- shared schema\\
- allowed features\\
- permitted link types\\
- compression modes

Stage 3---Dependency Exchange

Local universe announces:

I depend on: - remote\_node A - remote\_node B - chunk C

Stage 4---Partial Subgraph Import

Remote universe sends:\\
- minimal closure\\
- provenance fragments\\
- signature attestations

Stage 5---Reference Linking

Local universe creates:\\
- FederationEdge node linking local $\rightarrow$ remote

Stage 6---Causal Verification

Local universe checks:\\
- signatures\\
- invariants\\
- metadata\\
- rewrite rules

Stage 7---Finalization

Universe link established.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{8}
\tightlist
\item
  Federation Use Cases
\end{enumerate}

9.1 Multi-Repo Software Ecosystems

Thousands of DAGs with dependencies.

9.2 ML / LLM Training Pipelines

Link model evolution graphs across universes.

9.3 Scientific Workflow Federation

Massive cross-lab reproducible lineage.

9.4 C\ensuremath{\Omega}MPUTER: Simulation Inside Simulation

Nested universes.

9.5 Cross-Organizational Collaboration

Secure federation with limited permissions.

9.6 Distributed Multi-Agent Reasoning

Agents operating across multiple worlds.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{9}
\tightlist
\item
  Security
\end{enumerate}

Federation requires:\\
- signature verification\\
- universe policy negotiation\\
- key-based identity\\
- trust scoring\\
- denylist of hostile universes\\
- provenance validation

No universe should accept foreign truth without proof.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{10}
\tightlist
\item
  Why This Matters
\end{enumerate}

Because real computation isn't one universe.\\
It's an ecosystem of:\\
- different DAGs\\
- different policies\\
- different agents\\
- different semantic spaces\\
- different domains\\
- different teams

Federation is:\\
- package management\\
- dependency resolution\\
- distributed provenance\\
- inter-world communication\\
- DAG diplomacy\\
- computational cosmology

rolled into a single mechanism.

This RFC elevates JIT from ``universe'' to ``multiverse.''

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{11}
\tightlist
\item
  Status \& Next Steps
\end{enumerate}

We are ONE RFC away from\\
the JITOS Architecture Doc.

Possible next RFCs:\\
- RFC-0022---Visualization \& Rendering Protocol (JVP)\\
- RFC-0023---DAG Schema Evolution (DSE)\\
- RFC-0024---C\ensuremath{\Omega}MPUTER Fusion Layer\\
- RFC-0025---Distributed Scheduler / Agent Economy\\
- RFC-0026---Meta-Graph Interchange Format

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{\texorpdfstring{\textbf{C\ensuremath{\Omega}MPUTER {\textbullet}
JITOS}}{C\ensuremath{\Omega}MPUTER {\textbullet} JITOS}}\label{cux3c9mputer-jitos}

{\textcopyright} 2025 James Ross {\textbullet} \href{https://flyingrobots.dev}{Flying {\textbullet} Robots} All
Rights Reserved
