\section{RFC-0018}\label{rfc-0018}

JIT RFC-0018

Causal Garbage Isolation (CGI v1.0)

Entropy Management for an Immutable, Append-Only Universe

Status: Draft\\
Author: James Ross\\
Contributors: JIT Community\\
Requires:\\
- RFC-0001 Node Identity\\
- RFC-0002 DAG Invariants\\
- RFC-0003 SWS\\
- RFC-0005 Inversion Engine\\
- RFC-0009 Storage Tiering**\\
Start Date: 2025-11-29\\
Target Spec: JITOS v0.x**\\
License: TBD**

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Summary
\end{enumerate}

This RFC defines Causal Garbage Isolation (CGI)--- JIT's strategy for
managing unused or unreachable nodes in an append-only DAG\\
WITHOUT EVER DELETING ANYTHING.

Unlike traditional garbage collection (GC), which mutates or removes
data,\\
JIT uses a purely structural, causal, and reversible isolation system.

Causal Garbage Isolation:\\
- identifies unreachable subgraphs\\
- isolates them into ``dead zones''\\
- moves them to cold storage tiers\\
- preserves causal truth\\
- ensures deterministic replay\\
- prevents storage explosion\\
- enables long-term archiving\\
- allows optional re-animation of old universes

This turns ``garbage'' into dormant history, not trash.

Because the universe remembers everything.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  Motivation
\end{enumerate}

JIT is a:\\
- causal universe\\
- append-only timeline\\
- multi-agent simulation\\
- infinite DAG\\
- provenance-rich system

This means the DAG will grow forever.\\
And deleting nodes would:\\
- destroy reproducibility\\
- invalidate provenance\\
- break commits\\
- violate semantics\\
- erase realities\\
- collapse worldlines\\
- break physics

Deletion is unacceptable.\\
Mutation is forbidden.

Thus, the challenge:

How to manage infinite entropy without ever destroying truth?

Answer:

Causal Garbage Isolation.

Move unused nodes out of the ``active universe''\\
into frozen sectors of the graph-space\\
where they remain valid but inert.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  What Counts as Garbage? (Formally)
\end{enumerate}

A node N is garbage-like if:\\
- it is not reachable from any ref\\
- it is not reachable from any active SWS\\
- it is not part of any rewrite chain\\
- it is not referenced in any provenance\\
- it does not contribute to MH\\
- it is older than a configurable threshold\\
- it belongs to an abandoned branch\\
- it was created by an ephemeral agent\\
- it was a temporary result of an automated transformation

BUT:

A node is NEVER ``garbage.''\\
Nodes are simply less relevant.

Thus they are isolated, not deleted.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  The Isolation Process
\end{enumerate}

Isolation MUST follow this procedure:

Step 1---Identify unreachable nodes

Using graph traversal:

reachable = union(refs, SWS roots, rewrite parents)\\
all\_nodes - reachable = isolate\_candidates

Step 2---Verify they are causally dead

A node can only be isolated if:\\
- no future node references it\\
- no rewrite references it\\
- no provenance references it

Step 3---Move to Cold Tier

Nodes are moved from:

hot $\rightarrow$ warm $\rightarrow$ cold $\rightarrow$ deep-cold

Step 4---Mark with Isolation Flag

Index entry is updated:

isolated: true

This prevents:\\
- unnecessary rehydration\\
- unnecessary indexing\\
- unnecessary MH/FST projections

Step 5---Archive Subgraph

Nodes MAY be grouped into archive bundles for:\\
- backup\\
- remote cold storage\\
- long-term reproducibility\\
- scientific experiments

Step 6---Potential Reanimation

Users MAY request a subgraph be:

rehydrated $\rightarrow$ reintegrated $\rightarrow$ reattached

This ``resurrection'' is fully supported.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  Isolation Guarantees
\end{enumerate}

CGI MUST guarantee:

5.1 No Deletion

Nodes are never removed or destroyed.

5.2 DAG Stability

Isolation MUST never break invariants.

5.3 Deterministic Replay

Re-isolation MUST NOT alter future behavior.

5.4 No Mutation of Node Identity

Tier transitions preserve content.

5.5 Reversible Archiving

Nodes may be brought back.

5.6 Zero Impact on Truth

Isolation does not affect commit semantics.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{5}
\tightlist
\item
  Node States
\end{enumerate}

Nodes have four possible lifecycle states:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Hot
\end{enumerate}

Active, used by MH / SWS.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  Warm
\end{enumerate}

Compressed, mid-term history.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  Cold
\end{enumerate}

Long-term storage.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  Isolated Cold
\end{enumerate}

Detached from working DAG, preserved indefinitely.

Optionally:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  Deep-Cryo Cold
\end{enumerate}

For extremely old histories, stored offline.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{6}
\tightlist
\item
  Policies for Isolation
\end{enumerate}

JITD MUST allow:\\
- time-based isolation\\
- ref-based isolation\\
- age-based isolation\\
- branch abandonment isolation\\
- agent-scoped isolation\\
- SWS-scoped isolation\\
- storage-pressure-based isolation

Examples:\\
- ``isolate anything older than 90 days not reachable from a ref''\\
- ``isolate abandoned experimental branches''\\
- ``isolate LLM-generated ephemeral edits''\\
- ``only keep last 10K nodes in hot tier''

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{7}
\tightlist
\item
  Security Implications
\end{enumerate}

Isolation MUST NOT:\\
- bypass signing\\
- hide tampering\\
- erase provenance\\
- suppress audit data

Isolated nodes are still:\\
- queryable via JQL\\
- signed\\
- valid\\
- reversible

Just not active.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{8}
\tightlist
\item
  Why Causal Garbage Isolation Matters
\end{enumerate}

Because:

This is the ONLY GC model compatible with immutability, provenance,
reproducibility, AND infinite DAG growth.

It allows JIT to:\\
- run forever\\
- scale to billions of nodes\\
- support scientific workloads\\
- enable massive agent swarms\\
- maintain universal reproducibility\\
- archive entire universes\\
- perform deep causal forensics

JIT becomes:\\
- a universal archive\\
- a reproducible simulation environment\\
- a multi-agent OS\\
- a temporal database\\
- a causal memory engine

WITHOUT EVER THROWING ANYTHING AWAY.

This is how a universe manages entropy.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{9}
\tightlist
\item
  Status \& Next Steps
\end{enumerate}

The next foundational RFCs we can generate are:\\
- RFC-0019---SWS Memory Model\\
- RFC-0020---Security \& Permissions Model\\
- RFC-0021---Multi-Universe Federation Protocol (holy shit)\\
- RFC-0022---JIT Visualization Protocol\\
- RFC-0023---DAG Schema Evolution\\
- RFC-0024---C\ensuremath{\Omega}MPUTER Fusion Layer ({\bfseries ?})

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{\texorpdfstring{\textbf{C\ensuremath{\Omega}MPUTER {\textbullet}
JITOS}}{C\ensuremath{\Omega}MPUTER {\textbullet} JITOS}}\label{cux3c9mputer-jitos}

{\textcopyright} 2025 James Ross {\textbullet} \href{https://flyingrobots.dev}{Flying {\textbullet} Robots} All
Rights Reserved
