\section{RFC-0014}\label{rfc-0014}

JIT RFC-0014

Capability Negotiation \& Versioning (CNV v1.0)

Evolution Rules for a Causal Operating System

Status: Draft**\\
Author: James Ross\\
Contributors: JIT Community\\
Requires:\\
- RFC-0007 (RPC)\\
- RFC-0013 (ABI)**\\
Start Date: 2025-11-28**\\
Target Spec: JITOS v0.x**\\
License: TBD**

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Summary
\end{enumerate}

This RFC defines:\\
- how JITD announces capabilities\\
- how clients announce capabilities\\
- how versioning works\\
- how feature negotiation occurs\\
- how backward/forward compatibility is achieved\\
- how experimental features are isolated\\
- how the ecosystem evolves without breaking

This is the protocol for the future.

Every extension to JIT, every new RFC, every new operation MUST be
integrated through this mechanism.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  Motivation
\end{enumerate}

A causal OS must:\\
- evolve\\
- extend\\
- acquire features\\
- gain subsystems\\
- support new transports\\
- support new node types\\
- support new rewrite semantics\\
- remain stable

Traditional OSes solve this through:\\
- syscalls\\
- versioning\\
- feature detection\\
- flags

But JIT is distributed, agent-driven, and depends on:\\
- deterministic decoding\\
- binary integrity\\
- canonical encoding\\
- causal invariants\\
- compatibility across time and space

Therefore, capability negotiation is mandatory.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  Versioning Model
\end{enumerate}

JIT uses semantic versioning for kernel-level compatibility:

MAJOR.MINOR.PATCH

Encoded as:\\
- u16 in the ABI header\\
- u16 in handshake payload:\\
- MAJOR \textless\textless{} 8 \textbar{} MINOR\\
- PATCH is advisory only

MAJOR version bump:

Breaking change\\
(new invariants, new node formats)

MINOR version bump:

New features, backwards compatible\\
(new RPC ops, new node types)

PATCH version bump:

Bugfixes, documentation updates, clarifications

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  The Handshake
\end{enumerate}

When a client connects to JITD, it MUST send:

\{\\
``op'': ``handshake'',\\
``client\_version'': u16,\\
``capabilities'': {[}string{]},\\
``preferred\_compression'': string,\\
``supported\_transports'': {[}string{]}\\
\}

JITD responds:

\{\\
``status'': ``OK'',\\
``server\_version'': u16,\\
``capabilities'': {[}string{]},\\
``session\_id'': uuid,\\
``negotiated\_features'': {[}string{]},\\
``compression'': string,\\
``transport'': string\\
\}

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  Capability Strings
\end{enumerate}

Each capability MUST be a string like:

``jit.sws.v1''\\
``jit.collapse.v1''\\
``jit.sync.v1''\\
``jit.inversion.v1''\\
``jit.mh.v1''\\
``jit.storage.v1''\\
``jit.rpc.streaming''\\
``jit.security.signatures''\\
``jit.provenance.v1''\\
``jit.experimental.rewrite.v2''

Features MUST be:\\
- namespaced\\
- versioned\\
- descriptive\\
- future-proof

Clients \& servers MUST compare capabilities lexicographically.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{5}
\tightlist
\item
  Negotiation Rules
\end{enumerate}

6.1 Required Features

Kernel-level required features:

jit.dag\\
jit.wal\\
jit.refs\\
jit.abi\\
jit.rpc\\
jit.sws\\
jit.commit

If a client lacks a required feature $\rightarrow$\\
connection MUST fail.

{\bfseries ?}

6.2 Optional Features

Optional features may be:\\
- dropped\\
- accepted\\
- downgraded

Server chooses final set.

{\bfseries ?}

6.3 Experimental Features

EXPERIMENTAL features MUST begin with:

``jit.experimental.''

These:\\
- MUST NOT break determinism\\
- MUST NOT alter node identity\\
- MUST NOT alter invariants

Experimental features MUST NOT be enabled unless both parties explicitly
agree.

{\bfseries ?}

6.4 Conflicting Features

If two features conflict:\\
- server preference wins\\
- client may gracefully degrade\\
- or abort session

Conflicts MUST NOT lead to undefined behavior.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{6}
\tightlist
\item
  Future-Proofing
\end{enumerate}

The CNV model MUST support:\\
- new node types\\
- new storage tiers\\
- new SWS operations\\
- new transports\\
- new compression algorithms\\
- new cryptographic primitives

Without breaking old clients.

This is essential for JIT's long-term evolution.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{7}
\tightlist
\item
  Compatibility Guarantees
\end{enumerate}

8.1 Forward Compatibility

Old clients $\rightarrow$ new servers MUST still work,\\
but missing features are disabled.

8.2 Backward Compatibility

New clients $\rightarrow$ old servers MUST gracefully disable new features.

8.3 Deterministic Fallback

Fallback MUST NOT alter semantics of operations.

8.4 Universal Minimal Capability Set

All nodes/systems MUST support:\\
- CBOR canonicalization\\
- BLAKE3\\
- RFC-0001 node structure\\
- RFC-0002 invariants\\
- RFC-0013 ABI v1.0

This defines the minimum viable universe.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{8}
\tightlist
\item
  Security Considerations
\end{enumerate}

Capability negotiation MUST NOT:\\
- leak private server details\\
- allow downgrade attacks\\
- allow feature spoofing\\
- expose experimental features without consent

Signed handshakes (future RFC) will guarantee this.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{9}
\tightlist
\item
  Why CNV Matters
\end{enumerate}

Because:

Without it $\rightarrow$\\
extensions break ecosystems,\\
partial implementations fragment,\\
distributed systems desync,\\
and the kernel ossifies.

With CNV $\rightarrow$\\
JIT becomes an evolving universe\\
with infinite room for extensions,\\
experimentation,\\
academic theory,\\
industry adoption,\\
agent tooling,\\
and future OS-level patterns.

This is longevity.\\
This is stability.\\
This is how JIT survives the next 50 years.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{\texorpdfstring{\textbf{C\ensuremath{\Omega}MPUTER {\textbullet}
JITOS}}{C\ensuremath{\Omega}MPUTER {\textbullet} JITOS}}\label{cux3c9mputer-jitos}

{\textcopyright} 2025 James Ross {\textbullet} \href{https://flyingrobots.dev}{Flying {\textbullet} Robots} All
Rights Reserved
