\section{RFC-0020}\label{rfc-0020}

JIT RFC-0020

Security \& Permissions Model (SPM v1.0)

Root of Trust, Agent Authority, and Global Causal Integrity

Status: Draft\\
Author: James Ross\\
Contributors: JIT Community\\
Requires:\\
- RFC-0017 Agent Identity \& Signing\\
- RFC-0003 SWS\\
- RFC-0005 Inversion Engine\\
- RFC-0011 Distributed Sync\\
Start Date: 2025-11-29\\
Target Spec: JITOS v0.x\\
License: TBD

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Summary
\end{enumerate}

This RFC defines:\\
- the security boundaries\\
- permission rules\\
- access control\\
- commit authorization\\
- ref update policies\\
- SWS ownership model\\
- remote trust model\\
- sandboxing of LLM agents\\
- protections for DAG integrity\\
- replay protections\\
- signature requirements

Without this RFC, JIT is powerful but unsafe.\\
With this RFC, JIT becomes a secure kernel.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  Philosophy
\end{enumerate}

Security in JIT is grounded in:

Causal Integrity

Protect the DAG at all costs.\\
Truth is immutable.

Agent Accountability

All actions must be attributable.

Shadow Isolation

Agents cannot mutate the universe until commit.

Least Privilege

Shadows see only what they need.

Deterministic Enforcement

Security rules MUST be deterministic.

Distributed Trust

Remote peers MUST prove identity.

Semantic Safety

Agents MUST declare intent (provenance).

This model rejects:\\
- global mutable state\\
- root-level arbitrary writes\\
- unverified remote changes\\
- trust-by-default

We follow physics, not legacy computing.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  Security Domains
\end{enumerate}

JIT defines four major security domains:

3.1 DAG Domain (Truth)

Immutable. Sacred.\\
Only GITD updates it.\\
Only after deterministic collapse.\\
Protected by signatures and invariants.

3.2 SWS Domain (Shadow/Process)

Ephemeral but isolated.\\
Owned by an agent.\\
Sandboxed.\\
Limited authority.

3.3 Materialized Head (Human Projection)

Local representation.\\
User-facing.\\
Not authoritative.

3.4 Distributed Sync Domain

Remote peers, object stores, cross-machine negotiation.\\
Must be authenticated, validated, and replay-safe.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  Permission Model
\end{enumerate}

Every action MUST be explicitly authorized.

4.1 Commit Permissions

An agent MAY commit a given SWS only if:\\
- they own the SWS\\
- or have delegated authority

Collapse MUST verify:\\
- SWS.owner\\
- agent signature\\
- permission flags

Unauthorized collapse MUST fail.

{\bfseries ?}

4.2 Ref Update Permissions

Branch ref updates MUST require:\\
- agent signature\\
- authorization for that branch\\
- optional approval rules (multi-sig)

Tag ref updates MUST be forbidden unless:\\
- explicit override\\
- explicit permissions\\
- administrative mode

{\bfseries ?}

4.3 Node Creation Permissions

Any agent may propose overlay nodes\\
(because SWS allows experimentation),\\
BUT ONLY GITD:\\
- attaches them to DAG\\
- generates snapshot/rewrite nodes\\
- updates refs\\
- logs via WAL\\
- enforces invariants

Unauthorized node injection MUST fail.

{\bfseries ?}

4.4 Provenance Permissions

Any agent MAY write provenance nodes\\
as long as:\\
- agent identity is valid\\
- signature is valid\\
- metadata is in allowed schema\\
- size limits not exceeded\\
- SWS domain permits it

Provenance injection into the DAG DOES NOT alter truth,\\
but MUST be validated.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  Access Control (ACLs)
\end{enumerate}

ACL primitives include:

allow(agent, action, resource)\\
deny(agent, action, resource)\\
role(agent) = \{privileges\}\\
policy(branch) = \{rules\}

Resources:\\
- refs\\
- SWS\\
- nodes\\
- provenance categories\\
- remote sync endpoints\\
- MH projections

Actions:\\
- commit\\
- write-ref\\
- sync\\
- create-SWS\\
- attach-provenance\\
- read-node\\
- query-JQL

ACL rules MUST be deterministic.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{5}
\tightlist
\item
  Sandbox Model for Agents
\end{enumerate}

Agents MUST be sandboxed:

6.1 LLM Agents

LLMs MUST operate in:\\
- isolated SWS\\
- limited memory\\
- no direct DAG access\\
- no raw ref updates\\
- provenance enforcement\\
- signature enforcement\\
- rate-limited MP access

LLMs CAN propose changes.\\
GITD decides.

{\bfseries ?}

6.2 Human Agents

Humans may be granted:\\
- direct ref authority\\
- advanced collapse authority\\
- provenance override

They must still sign everything.

{\bfseries ?}

6.3 System Agents

CI, linkers, formatters, etc.\\
operate in low-permission SWS\\
with tightly restricted actions.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{6}
\tightlist
\item
  Distributed Sync Security
\end{enumerate}

Remote sync MUST enforce:\\
- signature validation\\
- trust policies\\
- denylist of malicious peers\\
- ref update restrictions\\
- causal invariant checks\\
- secure transport (TLS, QUIC)\\
- replay protection via logical timestamps

Remote nodes MUST NOT be accepted without validation.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{7}
\tightlist
\item
  Integrity Protections
\end{enumerate}

The system MUST:\\
- verify BLAKE3 hashes\\
- reject malicious DAG injections\\
- refuse history rewrites\\
- refuse corrupted rewrite nodes\\
- enforce invariants during replay\\
- ensure no unauthorized ref changes

If invariants fail $\rightarrow$\\
panic, quarantine, or safe-mode execution.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{8}
\tightlist
\item
  Multi-Signature Mode (Optional)
\end{enumerate}

Branches may require:\\
- 2-of-3 signatures\\
- 3-of-5 committee\\
- human-in-the-loop approvals\\
- CI + human pairing

This enables enterprise workflows.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{9}
\tightlist
\item
  Audit Trails
\end{enumerate}

All security-relevant operations MUST be logged:\\
- SWS creation/destruction\\
- commit events\\
- ref updates\\
- provenance\\
- agent identity delegation\\
- capability negotiation\\
- sync events\\
- failed or rejected collapses

Audit trails MUST be stored as DAG nodes or append-only logs for
verification.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{10}
\tightlist
\item
  Why Security Matters
\end{enumerate}

Because JIT is:\\
- the first OS for multi-agent systems\\
- the underlying physics for C\ensuremath{\Omega}MPUTER\\
- the substrate for agent-based software development\\
- the host of automated reasoning\\
- the truth layer for distributed systems\\
- the future backbone of reproducible computation

Without a solid security model:\\
- LLMs go rogue\\
- distributed sync becomes dangerous\\
- DAG poisoning attacks break truth\\
- collapse events can be forged\\
- humanity loses its audit trail

With this RFC?

JIT becomes secure enough to run the world.

{\bfseries ?}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{11}
\tightlist
\item
  Status \& Next Steps
\end{enumerate}

We have now completed the full safety layer of JITOS.

Possible next RFCs:\\
- RFC-0021---Multi-Universe Federation Protocol (HUGE)\\
- RFC-0022---Visualization \& Rendering Protocol\\
- RFC-0023---DAG Schema Evolution\\
- RFC-0024---C\ensuremath{\Omega}MPUTER Fusion Layer\\
- RFC-0025---Distributed Scheduler (Agent Economy)

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{\texorpdfstring{\textbf{C\ensuremath{\Omega}MPUTER {\textbullet}
JITOS}}{C\ensuremath{\Omega}MPUTER {\textbullet} JITOS}}\label{cux3c9mputer-jitos}

{\textcopyright} 2025 James Ross {\textbullet} \href{https://flyingrobots.dev}{Flying {\textbullet} Robots} All
Rights Reserved
