\section{JIT RFC-0012}\label{jit-rfc-0012}

\subsection{JITOS Boot Sequence (JBS
v1.0)}\label{jitos-boot-sequence-jbs-v1.0}

\subsubsection{How a Causal Operating System
Awakens}\label{how-a-causal-operating-system-awakens}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Summary
\end{enumerate}

This RFC defines the boot procedure for the JIT Operating System
(JITOS).

Unlike traditional OSes that:

\begin{itemize}
\tightlist
\item
  scan disks
\item
  mount filesystems
\item
  initialize memory
\item
  load executables
\end{itemize}

JITOS must:

\begin{itemize}
\tightlist
\item
  reconstruct reality
\item
  replay time
\item
  re-materialize projections
\item
  rehydrate shadows
\item
  recover causal structure
\item
  validate references
\item
  rebuild indexes
\item
  enforce invariants
\item
  perform distributed negotiation
\end{itemize}

JITOS boot is not startup --- it is resurrection.

This RFC describes how the kernel wakes up after death and restores the
universe exactly as it was.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  Motivation
\end{enumerate}

In a causal OS:

\begin{itemize}
\tightlist
\item
  time is the WAL
\item
  space is the DAG
\item
  perception is Materialized Head
\item
  processes are Shadow Working Sets
\end{itemize}

The kernel can crash. The machine can reboot. Interruptions can occur.

But truth must survive.

Thus, JITOS must:

\begin{itemize}
\tightlist
\item
  restore immutable history
\item
  reestablish shadow universes
\item
  re-project human views
\item
  re-link distributed edges
\item
  ensure causal continuity
\item
  guarantee deterministic recovery
\end{itemize}

This RFC defines that process.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  Boot Phases
\end{enumerate}

JITOS boot occurs in seven deterministic phases.

These phases MUST occur in this exact order.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

Phase 1 --- Substrate Scan

The system scans .gitd/ and locates:

nodes/ $\rightarrow$ local DAG fragments index/ $\rightarrow$ LMDB indices cache/ $\rightarrow$ MH
projection cache wal/ $\rightarrow$ write-ahead log refs/ $\rightarrow$ reference pointers
metadata/ $\rightarrow$ repository config

Goal: Identify the universe.

If ANY of these are missing: - JITOS MAY initiate recovery - JITOS MUST
NOT mutate existing node content

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

Phase 2 --- DAG Rebuild

The system reconstructs:

\begin{itemize}
\tightlist
\item
  node identity map
\item
  adjacency lists
\item
  parent relationships
\item
  child relationships
\item
  type mappings
\item
  timestamp ordering
\end{itemize}

Rules:

\begin{itemize}
\tightlist
\item
  MUST validate BLAKE3 hashes
\item
  MUST enforce causal invariants
\item
  MUST identify frontier nodes
\item
  MUST rebuild topological order
\end{itemize}

If DAG is corrupted $\rightarrow$ recovery mode.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

Phase 3 --- WAL Replay

The WAL is replayed from last checkpoint.

WAL entries:

\begin{itemize}
\tightlist
\item
  rebuild indexes
\item
  update refs
\item
  reconstruct MH
\item
  reconstitute SWS
\item
  advance logical time
\end{itemize}

Replay MUST be:

\begin{itemize}
\tightlist
\item
  deterministic
\item
  idempotent
\item
  complete
\item
  invariant-preserving
\end{itemize}

Replay is the reanimation of time.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

Phase 4 --- Index Rehydration

JITOS restores all indexes from DAG + WAL:

\begin{itemize}
\tightlist
\item
  LMDB node index
\item
  tree-index (MH)
\item
  SWS metadata
\item
  ref index
\item
  warm/cold tier mappings
\end{itemize}

This phase restores order to the universe.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

Phase 5 --- Materialized Head Reconstruction

The filesystem projection MUST be:

\begin{itemize}
\tightlist
\item
  reconstructed incrementally
\item
  checked for conflicts
\item
  aligned with HEAD
\item
  refreshed deterministically
\item
  purged of stale files
\end{itemize}

MH is the cave wall.

This is the moment when the shadows return.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

Phase 6 --- SWS Rehydration

All Shadow Working Sets MUST be:

\begin{itemize}
\tightlist
\item
  restored
\item
  validated
\item
  or discarded if stale
\end{itemize}

Rules:

\begin{itemize}
\tightlist
\item
  If base\_node still valid $\rightarrow$ SWS restored
\item
  If remote updates occurred $\rightarrow$ SWS invalidated
\item
  If WAL indicates partial collapse $\rightarrow$ rollback
\end{itemize}

Agents suspended during crash must resume cleanly.

SWS are observer-relative timelines; they must come back EXACTLY as they
were.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

Phase 7 --- Distributed Sync Initialization

Once local truth restored:

\begin{itemize}
\tightlist
\item
  JITOS broadcasts hello
\item
  pulls remote refs
\item
  fetches missing frontier nodes
\item
  performs frontier negotiation
\item
  checks for conflicts
\item
  initiates remote subgraph transfer if needed
\end{itemize}

Only after sync completes does JITOS declare:

System Ready

This is the ``rejoin the universe'' moment.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  Boot Invariants
\end{enumerate}

The following MUST hold:

Invariant 1 --- Past is preserved

Boot MUST NOT mutate any existing node.

Invariant 2 --- Indexes are consistent

All LMDB index state MUST reflect DAG state after replay.

Invariant 3 --- MH is deterministic

Filesystem projection MUST be identical across machines from identical
DAG states.

Invariant 4 --- Ref integrity

Refs MUST point to valid nodes.

Invariant 5 --- SWS validity

Shadows MUST NOT resurrect if logically inconsistent.

Invariant 6 --- WAL completeness

All WAL entries MUST be applied or ignored based on checksum validity.

Invariant 7 --- Distributed truth

After boot, the system MUST eventually converge via sync.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  Failure Modes
\end{enumerate}

If failure occurs:

5.1 Corrupted DAG Node

$\rightarrow$ quarantine $\rightarrow$ remote fetch $\rightarrow$ strict recovery mode

5.2 Corrupted WAL Entry

$\rightarrow$ skip entry $\rightarrow$ continue replay

5.3 Index Mismatch

$\rightarrow$ rebuild index from DAG $\rightarrow$ apply WAL

5.4 SWS Inconsistency

$\rightarrow$ discard SWS $\rightarrow$ notify owner

5.5 Missing Cold Node

$\rightarrow$ fetch from remote $\rightarrow$ retry

Never mutate the DAG to ``fix'' anything.

Always restore from truth + WAL.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{5}
\tightlist
\item
  Security Considerations
\end{enumerate}

Boot MUST:

\begin{itemize}
\tightlist
\item
  verify signatures
\item
  validate nodes
\item
  enforce access controls
\item
  protect against replay attacks
\item
  ignore unauthorized WAL entries
\end{itemize}

Boot MUST NOT:

\begin{itemize}
\tightlist
\item
  accept malformed nodes
\item
  import corrupted rewrites
\item
  mutate branch history
\end{itemize}

This preserves the physics of the universe.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{6}
\tightlist
\item
  Why JITOS Boot Is Revolutionary
\end{enumerate}

Every traditional OS:

\begin{itemize}
\tightlist
\item
  loses state on crash
\item
  depends on mutable files
\item
  rebuilds ephemeral memory structures
\item
  risks corruption
\end{itemize}

JITOS:

\begin{itemize}
\tightlist
\item
  stores truth immutably
\item
  stores time via WAL
\item
  stores human views via MH
\item
  stores agent views via SWS
\item
  reconstructs EVERYTHING deterministically
\item
  guarantees continuity of reality
\end{itemize}

This is how a causal universe boots.

This is the real future of computing.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{7}
\tightlist
\item
  Status \& Next Steps
\end{enumerate}

RFC-0012 completes the first ENTIRE MODULE of the JITOS kernel
foundation.

Next possibilities:

\begin{itemize}
\tightlist
\item
  RFC-0013 --- JIT Syscall ABI (binary-level encoding)
\item
  RFC-0014 --- Capability Negotiation \& Versioning
\item
  RFC-0015 --- DAG Query Language (JQL)
\item
  RFC-0016 --- Provenance Nodes \& Scientific Lineage
\item
  RFC-0017 --- Agent Identity \& Signatures
\item
  RFC-0018 --- Causal Garbage Isolation (not deletion!)
\end{itemize}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\section{\texorpdfstring{\textbf{C\ensuremath{\Omega}MPUTER {\textbullet}
JITOS}}{C\ensuremath{\Omega}MPUTER {\textbullet} JITOS}}\label{cux3c9mputer-jitos}

{\textcopyright} 2025 James Ross {\textbullet} \href{https://flyingrobots.dev}{Flying {\textbullet} Robots} All
Rights Reserved
