\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\fi
\usepackage{lmodern}
\usepackage{microtype}

\let\Bbbk\relax
\usepackage{newtxmath}
\usepackage{bm}
\let\mathbb\varmathbb
\let\openbox\relax

\usepackage{amsmath,amsthm,mathtools}
\usepackage{stmaryrd}
\DeclareFontFamily{U}{stmry}{}
\AtBeginDocument{%
  \DeclareFontShape{U}{stmry}{m}{n}{%
    <5> stmary5
    <5.5> stmary5
    <6> stmary6
    <7> stmary7
    <8> stmary8
    <9> stmary9
    <10> stmary10
    <10.5> stmary10
    <10.95> stmary10
    <12> stmary10
    <14.4> stmary10
    <17.28> stmary10
    <20.74> stmary10
    <24.88> stmary10
  }{}
}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,shapes.misc,positioning,calc,decorations.pathreplacing}
\usepackage{tikz-cd}
\usepackage{array}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage[nameinlink,capitalise]{cleveref}
\usepackage{enumitem}

\usepackage{titlesec}

\newcommand{\sectionbreak}{\clearpage}

\input{macros}

\title{Computational Holography\\[0.4em]
  \large Recursive Metagraphs, Deterministic Rewriting,\\
  and a Metric Geometry of Rulial Space}

\author{James Ross\\
  \small Independent Researcher\\
  \small \texttt{james@flyingrobots.dev}}

\date{\PaperDate}

\begin{document}

\hypersetup{pageanchor=false}
\pagenumbering{roman}
\input{titlepage}

\begin{abstract}
This paper develops a formal model of \emph{computational holography}: 
representing a computation so that its entire interior evolution 
is encoded on the boundary—specifically, the provenance payload of 
a single recursive metagraph edge. We build on double-pushout (DPO) graph 
rewriting in adhesive categories to define Recursive Metagraphs (RMGs), 
give a deterministic concurrent operational semantics, and prove 
tick-level confluence together with a two-plane commutation theorem. 
We also identify conditions for global confluence. Incorporating 
provenance payloads yields an information-complete, holographic 
encoding of computational history. Finally, we define an MDL-based 
\emph{rulial distance} between observers, relate RMG dynamics to multiway 
systems and the Ruliad, and outline how this structure supports the 
\AION{} \COMPUTER{}: a provenance-native computational model with 
deterministic tick semantics and a quasi-pseudometric 
geometry on observers.
\end{abstract}

\clearpage
\tableofcontents

\clearpage
\hypersetup{pageanchor=true}
\pagenumbering{arabic}

\input{assumptions}
\input{intro}
\input{rmg}
\input{dpo_rmg}
\input{determinism_confluence}
\input{holography}
\input{wormholes}
\input{rulial_distance}
\input{multiway_ruliad}
\input{ethics}
\input{discussion}

\bibliographystyle{alpha}
\bibliography{references}

\appendix
\section*{Author's Note}
This paper provides the formal backbone of the \AION{} \COMPUTER{}
project: a provenance-native machine model implemented in the Echo
engine.  The informal motivation---recursive structure in nature,
Romanesco cauliflowers, and all---is developed elsewhere; here we focus
on the mathematics.

% Hack: force arXiv to run 4 passes by mimicking LaTeX rerun warning
\typeout{get arXiv to do 4 passes: Label(s) may have changed. Rerun}
\end{document}
